Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-24 main.tex
MLL1 and KS1 proofs
moved some figs

\newcommand{\vgraphof}[1]{V_{\graphof{#1}}}
\newcommand{\gA}{\graph{A}}
\newcommand{\gB}{\graph{B}}
\newcommand{\gBone}{\graph{B}_1}
\newcommand{\gC}{\graph{C}}
\newcommand{\gG}{\graph{G}}
\newcommand{\gH}{\graph{H}}
\newcommand{\gN}{\graph{N}}

\newcommand{\bA}{\bgraph{A}}
\newcommand{\bB}{\bgraph{B}}
\newcommand{\bC}{\bgraph{C}}
\newcommand{\bG}{\bgraph{G}}
\newcommand{\bH}{\bgraph{H}}

@@

In this paper we advance the idea that combinatorial proofs can provide
a notion of proof identity for first-order logic. \emph{First-order
combinatorial proofs} were introduced by Hughes in~\cite{hughes:fopws}.
For example, a first-order combinatorial proof of Smullyan's 
``drinker paradox'' $\drinkerformula$ is shown above-right,
a homomorphism from a partially coloured graph to a labelled graph.
However, even though Hughes proves soundness and completeness, the 
proof is highly unsatisfactory: (1) the soundness argument is extremely
long, intricate and cumbersome, and (2) the completeness proof does not 
allow a syntactic proof to be read back from a combinatorial proof, i.e.,
completeness is not \emph{full}.\todo{cite something for full comp?}
A fundamental problem is that not all combinatorial
proofs can be obtained as translations of sequent calculus proofs.

@@

The fograph of drinker formula $\drinkerformula=\veedrinkerformula$:
%
\begin{center}\vspace{2ex}\begin{tikzpicture}
\drinkerbase
\end{tikzpicture}\vspace{1ex}\end{center}
%
%
Condensed combinatorial proof of drinker formula(s):
%
\begin{center}\vspace{3ex}\begin{math}
\begin{tikzpicture}
\drinkerInline
\end{tikzpicture}
\hspace{10ex}
\begin{tikzpicture}
\veedrinkerInline
\end{tikzpicture}\end{math}\end{center}

@@

Both $\veedrinkerformula$ and $\variantveedrinkerformula$ have the same
rectified fograph $\drinkergraph$, shown below-left.
%
\begin{center}\begin{math}
%
\drinkergraphpic
%
\hspace{8ex}
%
\drinkercotreepic
%
\hspace{8ex}
%
\drinkerbindinggraphpic
%
\end{math}\end{center}

Lifting diagrams:
%
\liftingdiagrams

\figdrinkerbifib Fig.\,\ref{fig:drinkerbifib} shows the drinker bifibration, binding fibration, and skeleton.

\leapfig Fig.\,\ref{fig:leap} shows a fonet with a unique dualizer, and its leap graph.

@@

Fix pairwise disjoint countable sets $\VAR=\set{x,y,z,\ldots}$ of variables,
$\FUN=\set{f,g,\ldots}$ of function symbols, and 
$\PRED=\set{p,q,\ldots}$ of predicate symbols. Each function symbol and
each predicate symbol has a finite arity. The grammars below generate
the set $\TERM$ of \bfit{terms}, denoted by $s,t,u,\ldots$, the set
$\ATOM$ of \bfit{atoms}, denoted by $a,b,c,\ldots$, and the set
$\FORM$ of \bfit{formulas}, denoted by $A,B,C,\ldots$:
\begin{equation*}
  \begin{array}{r@{~}l}
    t &\coloneqq~ x \mid f(t_1,\dots,t_n)
    \\[.5ex]
    a &\coloneqq~ \ttt\mid\fff\mid p(t_1,\dots,t_n) \mid \dual p(t_1,\dots,t_n)
    \\[.5ex]
    A &\coloneqq~ a \mid A\vlan A \mid A\vlor A \mid \exists x.A \mid \forall x.A
\end{array}
\end{equation*}
where the arity of $f$ and $p$ is $n$. We consider the truth constants
$\ttt$ (\emph{true}) and $\fff$
(\emph{false}) as additional atoms, and we consider all formulas in negation
normal form, where \bfit{negation} $(\dual\cdot)$ is defined on
atoms and formulas via De Morgan's laws:
\begin{equation*}
  \begin{array}{ccc}
    \dual{\dual a} = a
    &
  \dual\ttt =\fff &
  \dual{p(t_1,\ldots,t_n)} = \dual{p}(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual\fff = \ttt &
  \dual{\dual p(t_1,\ldots,t_n)} = p(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual{\exists x.A} = \forall x.\dual{A}
  &
  \dual{A \vlan B} = \dual{A} \ \vlor \ \dual{B} 
  \\[.5ex]
  &
  \dual{\forall x.A} = \exists x.\dual{A}
  &
%  \\[.5ex]
  \dual{A \vlor B} = \dual{A} \ \vlan \ \dual{B}
  \end{array}
\end{equation*}

@@

\bfit{Sequents}, denoted by $\Gamma,\Delta,\ldots$, are finite
multisets of formulas, written as lists, separated by comma. The
\bfit{corresponding formula} \todo{Case $n=0$!} of a sequent $\Gamma=A_1,A_2,\ldots,A_n$
is the disjunction of its formulas: $\form\Gamma=A_1\vlor
A_2\vlor\cdots\vlor A_n$. A sequent is \bfit{rectified} iff its
corresponding formula is.

@@

Modules in cographs correspond precisely to the subtrees of the
cotrees (the term-trees constructing the graph via $+$ and~$\times$).\todo{Cotree
must alternate $+$ and $\times$. Skip this comment completely?}

@@

We define a mapping $\graphof\cdot$ from formulas to (labelled)
graphs, inductively as follows:

For a formula $A$, we can define its
associated fograph $\graphof{A}$ inductively by:\todo{duplicate two sentences above}
\begin{equation*}
  \begin{array}{r@{\;}l@{\hskip4em}r@{\;}l}
    \graphof{a} &= \single a \rlap{\hskip1em(for any atom $a$)}  \\[.9ex]
    \graphof{A \cor B} &= \graphof{A} + \graphof{B} &
    \graphof{\exists x . A} &= \single x \times \graphof{A} \\[.9ex]
    \graphof{A \cand B} &= \graphof{A} \times \graphof{B} &
    \graphof{\forall x . A} &= \single x + \graphof{A}    
  \end{array}
\end{equation*}
where $\single a$ (resp. $\single x$) is a single-vertex graph whose vertex is 
    labelled by $a$ (resp. $x$).

@@

\begin{proof}
  That $\graphof A$ is a logical cograph follows immediately from the definition and Theorem~\ref{thm:cograph}.
  The fact that every binder of $\graphof A$ is legal can be proved by
  structural induction on $A$.
  %% :
  %% \begin{itemize}
  %%   \item Trivial if $A$ is an atom.
  %%   \item The induction step uses the following fact:
  %%         if $M$ is a module of $\gG$ and $M' \subseteq M$, then $M'$ is a
  %%         module of $\gG$ if and only if it is a module of $\gG[X]$.
  %% \end{itemize}
  %% \todo{}\lutz{maybe mention also fn 2-5 in Dominic's arxiv-paper}
\end{proof}
%%\juihsuan{I'm still looking for a simple proof}
Note that $\graphof A$ is not necessarily a fograph if $A$ is not
rectified. If $A = (\forall x. p(x)) \vlor (\forall x. q(x))$, then
$\graphof A = \single x \ \single p(x) \ \single x \ \single q(x)$,
the scope of each $x$-binder contains all the vertices, in particular,
the two $x$-binders. On the other hand, there are non-rectified
formulas which are translated to fographs by $\graphof\cdot$. For
example, in the graph of $(\exists x. p(x)) \vlor (\exists x. q(x))$,
both $x$-binders are legal, as they are not in each other's
scope:
%
$\edgepairpic{x}{px}\mkern9mu\edgepairpic{x}{qx}$.
%
For this reason, we call a formula
\bfit{clean} if it does not contain subformulas of the form $(\forall
x.A)\vlor(\forall x.B)$ or $(\exists x.A)\vlan(\exists x.B)$,
\todo{need the four cases as in clear definition in fopws? Consider
$(\forall x.A)\vlor(B\vlor (\forall x.C))$. But then Lemma 6 doesn't hold.
(fopws proves only that there's a surjection from clear formulas onto fographs,
whose kernel is the six formula equalities)} and no
$x$-quantified formula occurs as subformula of another $x$-quantified
formula. Then we have:

\begin{lemma}
  If $A$ is clean iff\/ $\graphof A$ is a fograph.
\end{lemma}

\begin{proof}
  Induction on $A$, using  Theorem~\ref{thm:cograph}.
\end{proof}

Even though for every formula $A$ we can obtain an
equivalent clean formula by simply renaming some bound variables, this
is not unique up to $\alpha$-conversion, as is the case for
rectified formulas.

@@

\begin{proof}
  By a straightforward induction on $A$.
\end{proof}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

A graph homomorphism $\phi\colon \tuple{\vG, \eG} \rightarrow
\tuple{\vH, \eH}$ is a \bfit{fibration} if for all $v \in \vG$ and
$w\phi(v) \in \eH$, there exists a unique $\tilde{w} \in \vG$ such
that $\tilde{w}v \in \eG$ and $\phi(\tilde{w}) = w$, and is a
\bfit{skew fibration} if for all $v \in \vG$ and $w\phi(v) \in \eH$
there exists $\tilde{w} \in \vG$ such that $\tilde{w}v \in \eG$ and
$\phi(\tilde{w})w \notin \eH$.  A directed graph homomorphism is a
\bfit{fibration} if for all $v \in \vG$ and $\pair{w,\phi(v)} \in
\eH$, there exists a unique $\tilde{w} \in \vG$ such that $\pair{\tilde{w},v}
\in \eG$ and $\phi(\tilde{w}) = w$.

@@

>>>>

\newcommand{\vgraphof}[1]{V_{\graphof{#1}}}
\newcommand{\gA}{\graph{A}}
\newcommand{\gB}{\graph{B}}
\newcommand{\gBone}{\graph{B}_1}
\newcommand{\gC}{\graph{C}}
\newcommand{\gD}{\graph{D}}
\newcommand{\gG}{\graph{G}}
\newcommand{\gH}{\graph{H}}
\newcommand{\gN}{\graph{N}}

\newcommand{\bA}{\bgraph{A}}
\newcommand{\bB}{\bgraph{B}}
\newcommand{\bC}{\bgraph{C}}
\newcommand{\bD}{\bgraph{D}}
\newcommand{\bG}{\bgraph{G}}
\newcommand{\bH}{\bgraph{H}}

@@

In this paper we advance the idea that combinatorial proofs can provide
a notion of proof identity for first-order logic. \emph{First-order
combinatorial proofs} were introduced by Hughes in~\cite{hughes:fopws}.
For example, a first-order combinatorial proof of Smullyan's 
``drinker paradox'' $\drinkerformula$ is shown above-right,
a homomorphism from a partially coloured graph to a labelled graph.
However, even though Hughes proves soundness and completeness, the 
proof is highly unsatisfactory: (1) the soundness argument is extremely
long, intricate and cumbersome, and (2) the completeness proof does not 
allow a syntactic proof to be read back from a combinatorial proof, i.e.,
completeness is not \emph{full}~\cite{abramsky:jagadeesan:94}.\todo{cite something for full comp?}\lutz{is that enough?}
A fundamental problem is that not all combinatorial
proofs can be obtained as translations of sequent calculus proofs.

@@

Condensed combinatorial proof of drinker formula(s):
%
\begin{center}\vspace{3ex}\begin{math}
\begin{tikzpicture}
\drinkerInline
\end{tikzpicture}
\hspace{10ex}
\begin{tikzpicture}
\veedrinkerInline
\end{tikzpicture}\end{math}\end{center}

@@

@@

Fix pairwise disjoint countable sets $\VAR=\set{x,y,z,\ldots}$ of variables,
$\FUN=\set{f,g,\ldots}$ of function symbols, and 
$\PRED=\set{p,q,\ldots}$ of predicate symbols. Each function symbol and
each predicate symbol has a finite arity. The grammars below generate
the set $\TERM$ of \bfit{terms}, denoted by $s,t,u,\ldots$, the set
$\ATOM$ of \bfit{atoms}, denoted by $a,b,c,\ldots$, and the set
$\FORM$ of \bfit{formulas}, denoted by $A,B,C,\ldots$:
\begin{equation*}
  \begin{array}{r@{~}l}
    t &\coloneqq~ x \mid f(t_1,\dots,t_n)
    \\[.5ex]
    a &\coloneqq~ \ttt\mid\fff\mid p(t_1,\dots,t_n) \mid \dual p(t_1,\dots,t_n)
    \\[.5ex]
    A &\coloneqq~ a \mid A\vlan A \mid A\vlor A \mid \exists x.A \mid \forall x.A
\end{array}
\end{equation*}
where the arity of $f$ and $p$ is $n$.
For better readability of often omit parentheses and write simply $ft_1,\dots,t_n$ or $pt_1,\dots,t_n$.
We consider the truth constants
$\ttt$ (\emph{true}) and $\fff$
(\emph{false}) as additional atoms, and we consider all formulas in negation
normal form, where \bfit{negation} $(\dual\cdot)$ is defined on
atoms and formulas via De Morgan's laws:
\begin{equation*}
  \begin{array}{ccc}
    \dual{\dual a} = a
    &
  \dual\ttt =\fff &
  \dual{p(t_1,\ldots,t_n)} = \dual{p}(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual\fff = \ttt &
  \dual{\dual p(t_1,\ldots,t_n)} = p(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual{\exists x.A} = \forall x.\dual{A}
  &
  \dual{A \vlan B} = \dual{A} \ \vlor \ \dual{B} 
  \\[.5ex]
  &
  \dual{\forall x.A} = \exists x.\dual{A}
  &
%  \\[.5ex]
  \dual{A \vlor B} = \dual{A} \ \vlan \ \dual{B}
  \end{array}
\end{equation*}
Then we write $A\implies B$ as abbreviation for $\dual A\vlor B$.

@@

\bfit{Sequents}, denoted by $\Gamma,\Delta,\ldots$, are finite
multisets of formulas, written as lists, separated by comma. The
\bfit{corresponding formula} of a (non-empty) sequent $\Gamma=A_1,A_2,\ldots,A_n$
is the disjunction of its formulas: $\form\Gamma=A_1\vlor
A_2\vlor\cdots\vlor A_n$. A sequent is \bfit{rectified} iff its
corresponding formula is.

@@

%% Modules in cographs correspond precisely to the subtrees of the
%% cotrees (the term-trees constructing the graph via $+$ and~$\times$).\todo{Cotree
%% must alternate $+$ and $\times$. Skip this comment completely?}

@@

We define a mapping $\graphof\cdot$ from formulas to (labelled)
graphs, inductively as follows:
%% For a formula $A$, we can define its
%% associated fograph $\graphof{A}$ inductively by:\todo{duplicate two sentences above}
\begin{equation*}
  \begin{array}{r@{\;}l@{\hskip4em}r@{\;}l}
    \graphof{a} &= \single a \rlap{\hskip1em(for any atom $a$)}  \\[.9ex]
    \graphof{A \cor B} &= \graphof{A} + \graphof{B} &
    \graphof{\exists x . A} &= \single x \times \graphof{A} \\[.9ex]
    \graphof{A \cand B} &= \graphof{A} \times \graphof{B} &
    \graphof{\forall x . A} &= \single x + \graphof{A}    
  \end{array}
\end{equation*}
where $\single a$ (resp. $\single x$) is a single-vertex graph whose vertex is 
labelled by $a$ (resp. $x$).

\begin{example}
  Below is the fograph of drinker formula $\drinkerformula=\veedrinkerformula$:
%
\begin{center}\vspace{2ex}\begin{tikzpicture}
\drinkerbase
\end{tikzpicture}\vspace{1ex}\end{center}
%
%
\end{example}

@@

\begin{proof}
  That $\graphof A$ is a logical cograph follows immediately from the definition and Theorem~\ref{thm:cograph}.
  The fact that every binder of $\graphof A$ is legal can be proved by
  structural induction on $A$.
  %% :
  %% \begin{itemize}
  %%   \item Trivial if $A$ is an atom.
  %%   \item The induction step uses the following fact:
  %%         if $M$ is a module of $\gG$ and $M' \subseteq M$, then $M'$ is a
  %%         module of $\gG$ if and only if it is a module of $\gG[X]$.
  %% \end{itemize}
  %% \todo{}\lutz{maybe mention also fn 2-5 in Dominic's arxiv-paper}
\end{proof}
%%\juihsuan{I'm still looking for a simple proof}

\begin{remark}
Note that $\graphof A$ is not necessarily a fograph if $A$ is not
rectified. If $A = (\forall x. p(x)) \vlor (\forall x. q(x))$, then
$\graphof A = \single x \ \single p(x) \ \single x \ \single q(x)$,
the scope of each $x$-binder contains all the vertices, in particular,
the two $x$-binders. On the other hand, there are non-rectified
formulas which are translated to fographs by $\graphof\cdot$. For
example, in the graph of $(\exists x. p(x)) \vlor (\exists x. q(x))$,
both $x$-binders are legal, as they are not in each other's
scope:
$\edgepairpic{x}{px}\mkern9mu\edgepairpic{x}{qx}$.
\end{remark}

%
%% For this reason, we call a formula
%% \bfit{clean} if it does not contain subformulas of the form $(\forall
%% x.A)\vlor(\forall x.B)$ or $(\exists x.A)\vlan(\exists x.B)$,
%% \todo{need the four cases as in clear definition in fopws? Consider
%% $(\forall x.A)\vlor(B\vlor (\forall x.C))$. But then Lemma 6 doesn't hold.
%% (fopws proves only that there's a surjection from clear formulas onto fographs,
%% whose kernel is the six formula equalities)} and no
%% $x$-quantified formula occurs as subformula of another $x$-quantified
%% formula. Then we have:

%% \begin{lemma}
%%   If $A$ is clean iff\/ $\graphof A$ is a fograph.
%% \end{lemma}

%% \begin{proof}
%%   Induction on $A$, using  Theorem~\ref{thm:cograph}.
%% \end{proof}

%% Even though for every formula $A$ we can obtain an
%% equivalent clean formula by simply renaming some bound variables, this
%% is not unique up to $\alpha$-conversion, as is the case for
%% rectified formulas.

@@

\begin{proof}
  By a straightforward induction on $A$.
\end{proof}

\begin{example}
  Both $\veedrinkerformula$ and $\variantveedrinkerformula$ (which are
  equivalent modulo $\fequ$) have the same (rectified) fograph
  $\drinkergraph$, shown below-left.
  %
  \begin{equation*}
      \drinkergraphpic
      %
      \hspace{8ex}
      %
      \drinkercotreepic
      %
      \hspace{8ex}
      %
      \drinkerbindinggraphpic
  \end{equation*}
  In the middle above we show the \emph{cotree} of the underlying
  cograph (illustrating the idea behind Theorem~\ref{thm:cograph}) and
  on the right above is its binding graph $\bD$.
\end{example}

@@

\begin{figure}[!t]
  \begin{center}
    \vspace{-1ex}
    \begin{tikzpicture}[graph]
      \begin{scope}[xshift=-2.2cm]\twolinkfograph\end{scope}
        \begin{scope}[xshift=2.2cm]\twolinkleapgraph\end{scope}
    \end{tikzpicture}\vspace{-4ex}
  \end{center}%
  \caption{\label{fig:leap}A fonet (left) with
    dualizer $\protect\twolinkassignment$
    and its
    leap graph (right).}
\end{figure}

Figure~\ref{fig:leap} shows a fonet with a unique dualizer, and its leap graph.

@@

A graph homomorphism $\phi\colon \tuple{\vG, \eG} \rightarrow
\tuple{\vH, \eH}$ is a \bfit{fibration} if for all $v \in \vG$ and
$w\phi(v) \in \eH$, there exists a unique $\tilde{w} \in \vG$ such
that $\tilde{w}v \in \eG$ and $\phi(\tilde{w}) = w$ (indicated on the left below), and is a
\bfit{skew fibration} if for all $v \in \vG$ and $w\phi(v) \in \eH$
there exists $\tilde{w} \in \vG$ such that $\tilde{w}v \in \eG$ and
$\phi(\tilde{w})w \notin \eH$ (indicated in the middle below).  A directed graph homomorphism is a
\bfit{fibration} if for all $v \in \vG$ and $\pair{w,\phi(v)} \in
\eH$, there exists a unique $\tilde{w} \in \vG$ such that $\pair{\tilde{w},v}
\in \eG$ and $\phi(\tilde{w}) = w$ (indicated o the right below).
%
\liftingdiagrams

@@

\begin{example}
  Below on the left is a skew bifibration, with its binding fibration
  in the middle. We often omit the labeling in the upper graph, as
  done on the right below.
  $$
  \scalebox{.9}{
  \begin{tikzpicture}[graph]\drinkerfiblabelledpair{x}{x}
  \end{tikzpicture}
  \quad
  \begin{tikzpicture}[graph]\drinkerbindingfiblabelled{x}{x}
  \end{tikzpicture}
  \quad
  \begin{tikzpicture}[graph]\drinkerfib
  \end{tikzpicture}}
  $$
\end{example}
